cardinalR: Generating interesting high-dimensional data structures

### 2025-09-19 paper-cardinalR.Rmd
Updated Trefoil section

To introduce missing or incomplete regions on the manifold, `gen_scurvehole(n, p)` (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:scurve) b"} else { "Figure \\@ref(fig:scurve-proj) b"})`) removes points in a localized region centered around the middle vertical section of the S-curve. Following a similar approach as `gen_cubehole(n, p, r_hole)`, all observations within a fixed radius ($\sqrt{0.3}$) of the anchor point are excluded, creating a hole in the manifold while preserving the overall S-curve structure.

@@

The Trefoil is a closed, nontrivial one-dimensional manifold embedded in $3\text{-}D$ or $4\text{-}D$ space (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:trefoil)"} else { "Figure \\@ref(fig:trefoil-proj)"})`). The trefoil features topological complexity in the form of self-overlaps, making it a valuable test case for evaluating the ability of NLDR methods to preserve global structure, loops, and embeddings in high-dimensional data. Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:trefoil-tb-html)', '\\@ref(tab:trefoil-tb-pdf)'))` summarizes these functions.

@@

#### `gen_trefoil4d()`

The function `gen_trefoil4d()` generates a $4\text{-}D$ trefoil knot using two angular parameters, $\theta$ and $\phi$. The knot is constructed on the $3$-sphere ($S^3 \subset \mathbb{R}^4$), with a band of thickness around the knot path, controlled by `steps` (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:trefoil) a"} else { "Figure \\@ref(fig:trefoil-proj) a"})`).

The coordinates of the $4\text{-}D$ trefoil knot are defined as $X_1 = \cos(\theta) \cos(\phi)$, $X_2 = \cos(\theta) \sin(\phi)$, $X_3 = \sin(\theta) \cos(1.5 \phi)$, and $X_4 = \sin(\theta) \sin(1.5 \phi)$, where $\theta$ and $\phi$ are angular parameters that trace out the knot's structure on the $3$-sphere $S^3 \subset \mathbb{R}^4$.

The number of `theta` and `phi` values is determined by the `steps` and `n` arguments, respectively. Additional independent noise dimensions can be added via `gen_noisedims()` to extend the structure into $p > 4$ dimensions.

@@

#### `gen_trefoil3d()`

The function `gen_trefoil3d()` generates a $3\text{-}D$ stereographic projection of the $4\text{-}D$ trefoil knot by mapping each point $(X_1, X_2, X_3, X_4) \in \mathbb{R}^4$ to $(X_1', X_2', X_3') \in \mathbb{R}^3$ using the transformation $X_1' = X_1 / (1 - X_4)$, $X_2' = X_2 / (1 - X_4)$, and $X_3' = X_3 / (1 - X_4)$, while excluding points where $X_4 = 1$ to avoid division by zero (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:trefoil) b"} else { "Figure \\@ref(fig:trefoil-proj) b"})`).

@@

As with the $4\text{-}D$ version, this function supports optional noise dimensions to embed the projected knot into higher dimensions.

>>>>

To introduce missing or incomplete regions on the manifold, `gen_scurvehole(n, p)` (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:scurve) b"} else { "Figure \\@ref(fig:scurve-proj) b"})`) removes points in a localized region centered around the middle vertical section of the S-curve. Following a similar approach as `gen_cubehole()`, all observations within a fixed radius ($\sqrt{0.3}$) of the anchor point are excluded, creating a hole in the manifold while preserving the overall S-curve structure.

@@

The Trefoil is a closed, nontrivial one-dimensional manifold embedded in $3\text{-}D$ or $4\text{-}D$ space (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:trefoil)"} else { "Figure \\@ref(fig:trefoil-proj)"})`). The trefoil features topological complexity in the form of self-overlaps, making it a valuable test case for evaluating the ability of non-linear dimension reduction methods to preserve global structure, loops, and embeddings in high-dimensional data. Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:trefoil-tb-html)', '\\@ref(tab:trefoil-tb-pdf)'))` summarizes these functions.

@@

For the $4\text{-}D$ trefoil knot, the function `gen_trefoil4d(n, p, steps)` generates the structure on the $3$-sphere ($S^3 \subset \mathbb{R}^4$) using two angular parameters, $\theta$ and $\phi$. A band of thickness around the knot path is controlled by the `steps` argument, while the number of $\theta$ and $\phi$ values is determined by the `steps` and `n` arguments, respectively (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:trefoil) a"} else { "Figure \\@ref(fig:trefoil-proj) a"})`). The coordinates are defined as $X_1 = \cos(\theta) \cos(\phi)$, $X_2 = \cos(\theta) \sin(\phi)$, $X_3 = \sin(\theta) \cos(1.5 \phi)$, and $X_4 = \sin(\theta) \sin(1.5 \phi)$, where $\theta$ and $\phi$ trace the knot’s path. For $p > 4$, the quadratic structure is embedded into higher dimensions by appending additional noise dimensions.

@@

For the $3\text{-}D$ stereographic projection, `gen_trefoil3d(n, p, steps)` maps each point $(X_1, X_2, X_3, X_4) \in \mathbb{R}^4$ to $(X_1', X_2', X_3') \in \mathbb{R}^3$ using $X_1' = X_1 / (1 - X_4)$, $X_2' = X_2 / (1 - X_4)$, and $X_3' = X_3 / (1 - X_4)$, excluding points where $X_4 = 1$ to avoid division by zero (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:trefoil) b"} else { "Figure \\@ref(fig:trefoil-proj) b"})`). As with the $4\text{-}D$ case, optional noise dimensions can be added to embed the knot into higher-dimensional spaces.

@@

### 2025-09-19 paper-cardinalR.tex
Updated Trefoil section

To introduce missing or incomplete regions on the manifold, \texttt{gen\_scurvehole(n,\ p)} (Figure \ref{fig:scurve-proj} b) removes points in a localized region centered around the middle vertical section of the S-curve. Following a similar approach as \texttt{gen\_cubehole(n,\ p,\ r\_hole)}, all observations within a fixed radius (\(\sqrt{0.3}\)) of the anchor point are excluded, creating a hole in the manifold while preserving the overall S-curve structure.

@@

The Trefoil is a closed, nontrivial one-dimensional manifold embedded in \(3\text{-}D\) or \(4\text{-}D\) space (Figure \ref{fig:trefoil-proj}). The trefoil features topological complexity in the form of self-overlaps, making it a valuable test case for evaluating the ability of NLDR methods to preserve global structure, loops, and embeddings in high-dimensional data. Table \ref{tab:trefoil-tb-pdf} summarizes these functions.

@@

\paragraph{\texorpdfstring{\texttt{gen\_trefoil4d()}}{gen\_trefoil4d()}}\label{gen_trefoil4d}

The function \texttt{gen\_trefoil4d()} generates a \(4\text{-}D\) trefoil knot using two angular parameters, \(\theta\) and \(\phi\). The knot is constructed on the \(3\)-sphere (\(S^3 \subset \mathbb{R}^4\)), with a band of thickness around the knot path, controlled by \texttt{steps} (Figure \ref{fig:trefoil-proj} a).

The coordinates of the \(4\text{-}D\) trefoil knot are defined as \(X_1 = \cos(\theta) \cos(\phi)\), \(X_2 = \cos(\theta) \sin(\phi)\), \(X_3 = \sin(\theta) \cos(1.5 \phi)\), and \(X_4 = \sin(\theta) \sin(1.5 \phi)\), where \(\theta\) and \(\phi\) are angular parameters that trace out the knot's structure on the \(3\)-sphere \(S^3 \subset \mathbb{R}^4\).

The number of \texttt{theta} and \texttt{phi} values is determined by the \texttt{steps} and \texttt{n} arguments, respectively. Additional independent noise dimensions can be added via \texttt{gen\_noisedims()} to extend the structure into \(p > 4\) dimensions.

@@

\paragraph{\texorpdfstring{\texttt{gen\_trefoil3d()}}{gen\_trefoil3d()}}\label{gen_trefoil3d}

The function \texttt{gen\_trefoil3d()} generates a \(3\text{-}D\) stereographic projection of the \(4\text{-}D\) trefoil knot by mapping each point \((X_1, X_2, X_3, X_4) \in \mathbb{R}^4\) to \((X_1', X_2', X_3') \in \mathbb{R}^3\) using the transformation \(X_1' = X_1 / (1 - X_4)\), \(X_2' = X_2 / (1 - X_4)\), and \(X_3' = X_3 / (1 - X_4)\), while excluding points where \(X_4 = 1\) to avoid division by zero (Figure \ref{fig:trefoil-proj} b).

@@

As with the \(4\text{-}D\) version, this function supports optional noise dimensions to embed the projected knot into higher dimensions.

>>>>

To introduce missing or incomplete regions on the manifold, \texttt{gen\_scurvehole(n,\ p)} (Figure \ref{fig:scurve-proj} b) removes points in a localized region centered around the middle vertical section of the S-curve. Following a similar approach as \texttt{gen\_cubehole()}, all observations within a fixed radius (\(\sqrt{0.3}\)) of the anchor point are excluded, creating a hole in the manifold while preserving the overall S-curve structure.

@@

The Trefoil is a closed, nontrivial one-dimensional manifold embedded in \(3\text{-}D\) or \(4\text{-}D\) space (Figure \ref{fig:trefoil-proj}). The trefoil features topological complexity in the form of self-overlaps, making it a valuable test case for evaluating the ability of non-linear dimension reduction methods to preserve global structure, loops, and embeddings in high-dimensional data. Table \ref{tab:trefoil-tb-pdf} summarizes these functions.

@@

For the \(4\text{-}D\) trefoil knot, the function \texttt{gen\_trefoil4d(n,\ p,\ steps)} generates the structure on the \(3\)-sphere (\(S^3 \subset \mathbb{R}^4\)) using two angular parameters, \(\theta\) and \(\phi\). A band of thickness around the knot path is controlled by the \texttt{steps} argument, while the number of \(\theta\) and \(\phi\) values is determined by the \texttt{steps} and \texttt{n} arguments, respectively (Figure \ref{fig:trefoil-proj} a). The coordinates are defined as \(X_1 = \cos(\theta) \cos(\phi)\), \(X_2 = \cos(\theta) \sin(\phi)\), \(X_3 = \sin(\theta) \cos(1.5 \phi)\), and \(X_4 = \sin(\theta) \sin(1.5 \phi)\), where \(\theta\) and \(\phi\) trace the knot's path. For \(p > 4\), the quadratic structure is embedded into higher dimensions by appending additional noise dimensions.

@@

For the \(3\text{-}D\) stereographic projection, \texttt{gen\_trefoil3d(n,\ p,\ steps)} maps each point \((X_1, X_2, X_3, X_4) \in \mathbb{R}^4\) to \((X_1', X_2', X_3') \in \mathbb{R}^3\) using \(X_1' = X_1 / (1 - X_4)\), \(X_2' = X_2 / (1 - X_4)\), and \(X_3' = X_3 / (1 - X_4)\), excluding points where \(X_4 = 1\) to avoid division by zero (Figure \ref{fig:trefoil-proj} b). As with the \(4\text{-}D\) case, optional noise dimensions can be added to embed the knot into higher-dimensional spaces.

@@
